---
title: "Case Study: Clustering"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### Dataset: IRIS
* * *

* Load the dataset

```{r}
data(iris)
summary(iris)
```
* Remove the values from the **species** columns
```{r}
iris_species_unk = iris
iris_species_unk$Species = NULL
```

### K-Means
* * *

* The ``nstart`` allows to run different random starting assigments and to select the one with the lowest within cluster variation
* Ensure reproducibility by setting the seed
* Assume that *K* = 3

```{r}
# kmeans(x, centers, ...)
set.seed(20)
km_clusters = kmeans(iris_species_unk[], centers = 3, nstart = 20)
str(km_clusters)
```

* Compare the clusters with the species
```{r}
table(km_clusters$cluster, iris$Species)
```

* Plot data samples in clusters
```{r}
plot(iris_species_unk$Sepal.Length, iris_species_unk$Sepal.Width, col=km_clusters$cluster, xlab = 'Sepal Lenght', ylab = 'Sepal Width')
points(km_clusters$centers[,c('Sepal.Length', 'Sepal.Width')], col=1:3, pch=8, cex=2)
```
```{r}
library(ggplot2)
irisClusters = as.factor(km_clusters$cluster)
ggplot(iris_species_unk, aes(Petal.Length, Petal.Width, color = irisClusters)) + geom_point()
```

#### Optimal **K** computation
Model selection criteria:

* **AIC** (Akaike Information Criterion)
* **BIC** (Bayesian Information Criterion)

```{r}
aic_bic = function(fit){
  # Number of features #
  m = ncol(fit$centers)
  # Number of observations #
  n = length(fit$cluster)
  # Number of clusters, i.e. k #
  k = nrow(fit$centers)
  # Total within-cluster sum of squares
  D = fit$tot.withinss
  return(c(D + 2*m*k, D + log(n)*m*k))
}
```
* Which are the **AIC** and **BIC** values?
```{r}
values = aic_bic(km_clusters)
names(values) = c('AIC', 'BIC')
print(values)
```
* Pick the model with the lowest **BIC** or **AIC**
* Check values of **K** between 3 to 40
```{r}
#cat("K", "\t", "AIC", "\t\t", "BIC", "\n")
lowest_bic = lowest_aic = 1000
best_k_bic = best_k_aic = 0
for (k in 3:40) {
  aic_bic_k = aic_bic(kmeans(iris_species_unk, k))
  current_aic = aic_bic_k[1]
  current_bic = aic_bic_k[2]
  if (current_aic < lowest_aic) {
    lowest_aic = current_aic
    best_k_aic = k
  }
  if (current_bic < lowest_bic) {
    lowest_bic = current_bic
    best_k_bic = k
  }
  # cat(k, '\t', bic_aic_k[1], '\t', bic_aic_k[2], '\n')
}
```

```{r}
cat('Best K according to AIC: ', best_k_aic, '-- BIC: ', lowest_aic, '\n')
```
```{r}
cat('Best K according to BIC: ', best_k_bic, '-- BIC: ', lowest_bic, '\n')
```

##### Activity:
Apply the same analysis as before to the following dataset:

* ``install.packages('rattle.data')``
* ``data("wine", package = 'rattle.data')``

### Hierarchical Clustering
* * *
* Take a sample from the IRIS dataset:
```{r}
idx = sample(1:dim(iris)[1], 40)
iris_sample = iris[idx,]
iris_sample$Species = NULL
```
* Create the clusters
```{r}
hc_clusters = hclust(dist(iris_sample), method="ave")
str(hc_clusters)
```
* Plot the result
```{r}
plot(hc_clusters, hang = -1, labels=iris$Species[idx])
```
