---
title: "Case Study: Link Analysis"
output:
  html_document:
    df_print: paged
    #code_folding: hide
  pdf_document: default
---

```{r, echo=FALSE}
rm(list=ls())
```
___
### Objectives
* Explore a dataset on flight routes
* Plot the graph that shows the links between the airports (how the airports are connected)
* Analize differtent link analysis properties such as outdegree, betweenness and PageRank

### Flight Routes Dataset
* Download from: ```ir.inf.usi.ch/da-datasets/flights.zip``` and load it
```{r, echo=TRUE}
airports = read.csv("airports.dat", header=T)
routes = read.csv("routes.dat", header=T)
```
* Explore the data:
```{r}
nrow(airports)
head(airports)
summary(airports)

nrow(routes)
head(routes)
summary(routes)
```
* Select only airports located in United_Kingdom
```{r}
airports_United_Kingdom = airports[airports$Country == 'United Kingdom',]
```
* Filter out airports that don't have a specified *IATA* code
* Keep only the **City**, **IATA**, **Latitude**, **Longitude** columns in ``airports_United_Kingdom``
```{r}
airports_United_Kingdom = airports_United_Kingdom[airports_United_Kingdom$IATA != '', 
                                c('City', 'IATA', 'Latitude', 'Longitude')]
```
* Keep only the Italian routes
```{r}
iata_italian_codes = airports_United_Kingdom$IATA
routes_United_Kingdom <- routes[(routes$SourceAirport %in% iata_italian_codes) & 
                        (routes$DestinationAirport %in% iata_italian_codes),
                       c('SourceAirport','DestinationAirport')]
```
* How many airports there are in United_Kingdom?
```{r}
nrow(airports_United_Kingdom)
```
How many routes there are in United_Kingdom?
```{r}
nrow(routes_United_Kingdom)
```
###Plot the Flight Routes in United_Kingdom
* Associate the geographical coordinates of the airports of departure and arrival for each flying route
* Create a function to get the coordinates of an airport, given its **IATA** code:
* Function name: ```getCoordinates_United_Kingdom```
* Takes one argument: ```iata_codes```
* Gets the ```Longitude```, ```Latitude``` of the given ```iata_codes``` using ```airports_United_Kingdom``` variable
* Returns a data frame that contains the ```Longitude```, ```Latitude``` of the given iata

```{r}
getCoordinates_United_Kingdom = function(iata_code) {
  codefilter = airports_United_Kingdom$IATA == iata_code
  Longitude = airports_United_Kingdom[codefilter, 'Longitude']
  Latitude = airports_United_Kingdom[codefilter, 'Latitude']
  return (data.frame(Longitude, Latitude))
}
```

* Create departure and arrival locations based on their ```Longitude``` and ```Latitude```
* In order to do so, apply the ```getCoordinates_United_Kingdom()``` function to all the source and destination airports of United_Kingdom

```{r}
departure_locations = lapply(as.character(routes_United_Kingdom$SourceAirport), 
                             getCoordinates_United_Kingdom)
departure_locations = do.call('rbind', departure_locations)
departure_locations$group = 1 : dim(departure_locations)[1]

arrival_locations = lapply(as.character(routes_United_Kingdom$DestinationAirport),
                            getCoordinates_United_Kingdom)
arrival_locations = do.call('rbind', arrival_locations)
arrival_locations$group = 1 : dim(arrival_locations)[1]

routes_United_Kingdom_coordinates = rbind(departure_locations, arrival_locations)
```

* Install ```ggmap```
```{r, message=FALSE, warning=FALSE}
#install.packages('ggmap')
library(ggmap)
library(ggplot2)
```

* Copy and paste the function provided in the ```ggmap_extraFunctions.R``` file and create the plot

```{r, eval=FALSE, message=FALSE}
map_United_Kingdom = ggmap(get_googlemap(center = 'United Kingdom', zoom = 6,
                                maptype = 'roadmap', color = 'bw'),
                  extent='device') +
  geom_path(data = routes_United_Kingdom_coordinates,
            aes (x = Longitude , y = Latitude , group=group),
            size = 0.4 , alpha = I(0.05) , color = 'red', na.rm=TRUE) +
  geom_point(data = airports_United_Kingdom,
             aes (x = Longitude , y = Latitude),
             color = '#1565C0' , alpha = I(0.3), na.rm=TRUE)
print(map_United_Kingdom)
```

##### Rank the airports according to some metrics (outdegree, betweeness, PageRank)
* **Outdgree**: Number of edges outgoing from a vertex or node
* Create a function ```outdegree_United_Kingdom```
* The function should return the outdegree of an airport given its IATA code
* Consider only routes in United_Kingdom (```routes_United_Kingdom```)
* Store the outdegree of the Italian airports in the variable airports_United_Kingdom$Outdegree

```{r}
outdegree_United_Kingdom = function(iata) {
  outdegree = sum(routes_United_Kingdom$SourceAirport == iata, na.rm=T) 
  return (outdegree)
}
iata_unique = as.character(unique(airports_United_Kingdom$IATA))
airports_United_Kingdom$Outdegree = do.call('rbind', lapply(iata_unique, outdegree_United_Kingdom))
```

* Copy and paste the function provided in the ```ggmap_extraFunctions.R``` file and create the plot
```{r}
map_United_Kingdom_outdegree = ggmap(get_googlemap(center = 'United Kingdom',zoom = 6,
                                          maptype = 'roadmap',
                                          color = 'bw' ), 
                            extent = 'device') +
  geom_path (data = routes_United_Kingdom_coordinates,
             aes (x = Longitude, y = Latitude, group = group),
             size = 0.4, alpha = I(0.05), color = 'blue', na.rm=TRUE) +
  geom_point (data = airports_United_Kingdom,
              aes ( x = Longitude , y = Latitude, size = Outdegree),
              color = 'blue' , alpha = I(0.6), na.rm=TRUE)

print(map_United_Kingdom_outdegree)
```

* **Betweeness**: number of times that a node serves as a bridge along the shortest path between two other nodes
* Install the ```igraph``` library which contains network analysis and visualization functions
```{r}
#install.packages('igraph')
library('igraph')
```
* Build a graph where each airport is a vertex and each route is an edge
```{r}
graph_United_Kingdom_vertices = as.character(unique(airports_United_Kingdom$IATA))
graph_United_Kingdom = graph.empty() + vertices(graph_United_Kingdom_vertices)
graph_United_Kingdom = graph_United_Kingdom + 
  edges(as.matrix(routes_United_Kingdom[, c("SourceAirport", "DestinationAirport")]), directed=T)
```
* Compute the betweenness for each airport
```{r}
United_Kingdom_betweenness = betweenness(graph_United_Kingdom, v=V(graph_United_Kingdom), directed=T)
United_Kingdom_betweenness = as.matrix(United_Kingdom_betweenness )
airports_United_Kingdom = merge(airports_United_Kingdom, United_Kingdom_betweenness, by.x="IATA", by.y="row.names")
colnames(airports_United_Kingdom)[6] = "Betweenness"
```
* Copy and paste the function provided in the ```ggmap_extraFunctions.R``` file and create the plot
```{r}
map_United_Kingdom_betweenness = (ggmap(get_googlemap(center = 'United_Kingdom', zoom = 6, 
                                             maptype = 'roadmap', color = 'bw' ), 
                               extent = 'device') +
                           geom_path(data = routes_United_Kingdom_coordinates,
                                     aes ( x = Longitude , y = Latitude , group = group ),
                                     size = 0.1 , alpha = I(0.05) , color = '#e1dfdf', na.rm=TRUE ) +
                           geom_point (data = airports_United_Kingdom , 
                                         aes(x = Longitude, y = Latitude,
                                             colour = as.vector(Betweenness), size = Outdegree),
                                       alpha = I(0.8), na.rm=TRUE ) +
                           guides(colour=guide_colourbar("Betweenness")) +
                           scale_color_gradientn(guide="colourbar", na.value="black", colours=rainbow(2)))

print(map_United_Kingdom_betweenness)
```






